A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2017-01-30 rigid_bodies.tex
first modification

\section{Introduction}

Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations. This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.

@@

\subsection{Kinematics}
\label{sec:kinematics}

Consider a rigid body composed of $n_p$ individual particles, with each particle $j$ having mass $M_j$ and being located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
\begin{equation*}
\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
\end{equation*}
where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body. The position of each particle $j$ in a space-fixed frame of reference, whose origin coincides with the body's center of mass, is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$. In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
\begin{equation*}
{\mt I}^\ast = \sum_{j=1}^N M_j \begin{smallarray}{ccc}
{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
-{\delta_j}_y {\delta_j}_x & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z \\
-{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
\end{smallarray}.
\end{equation*}

For every non-collinear rigid body (for conciseness, the only case considered here), there exists a particular frame of reference in which its inertia tensor is diagonal. The formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\cite{Goldstein2002}
\begin{equation*}
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \begin{smallarray}{ccc}
I_1 &   0 &   0 \\
0 & I_2 &   0 \\
0 &   0 & I_3 \\
\end{smallarray},
\end{equation*}
where $I_1$, $I_2$, and $I_3$ (the eigenvalues of ${\mt I}^\ast$) are the principal moments of inertia and the columns of $\tr{\mt A}$ (the corresponding eigenvectors) are the principal axes of rotation. With no loss of generality, we can consider that $I_1 \geq I_2 \geq I_3$. Algorithms for obtaining $\mt A$ from $\mt I^\ast$ can be found in the literature (e.g. Ref.~\onlinecite{Kopp2008}). Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix. By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that the result of $\mt A \tr{\dot{\mt A}}$ is skew-symmetric. The angular velocity of the body expressed in the body-fixed frame, $\vt \omega$, is related to this product by $\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}}$,\cite{Haug1989} where $\mt S(\cdot)$ is the skew-symmetric operator defined in Appendix \ref{sec:auxiliary_math}  (Eq.~\ref{eq:operator_S}).

An arbitrary rotation can be described using three independent parameters. One can use Euler angles, for instance, but mathematical singularities arise when the equations of motion are described in such terms. Instead, we consider an alternative representation, in which the orientation of a rigid body is given by four parameters $q_1$, $q_2$, $q_3$, and $q_4$, often referred to as Euler parameters, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ and satisfy the relation\cite{Goldstein2002}
\begin{equation}
\label{eq:norm_eq_1}
\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1.
\end{equation}

Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain. Methods are available for computing the unit quaternion related to a given rotation matrix (e.g. Ref.~\onlinecite{Shepperd1978}). Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray}.
\end{equation*}

Some important properties of these matrices (or transposes thereof) have been reported in the literature\cite{Haug1989, Shuster1993, Dichmann1999, Ravishankar2004, Nielsen2012} and are revisited with our particular notation in Appendix~\ref{sec:auxiliary_math}. The most remarkable feature is that
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation}

Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$. It is explained in Appendix~\ref{sec:auxiliary_math} that premultiplication of a vector with either $\mt B$ or $\mt C$ transforms it into a quaternion that lies in the hyperplane orthogonal to $\vt q$ and that both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly. Therefore, there exists an intermediary frame of reference, which is fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body. Its connection to the body-fixed and space-fixed frames are summarized as follows:
\[
\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\mt B}_{\tr{\mt B}}}
\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\tr{\mt C}}_{\mt C}}
\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
\]

Hence, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is half the quaternion-fixed version of the angular velocity vector. Besides, it guarantees that the norm of $\vt q$ remains unchanged because Eq.~\ref{eq:diff_qTq} is identically satisfied.

To conclude this section, we discuss on the possibility of switching between alternative space-fixed frames when describing the motion of a rigid body. This will be useful when we deal with torque-free rotations (Sec.~\ref{sec:free_rotation}). Consider that a body-fixed vector $\vt d$ is simultaneously observed as $\vt \delta = \tr{\mt A\!}(\vt q){\vt d}$ in a space-fixed frame $\mathcal S_q$ and as $\vt \xi = \tr{\mt A\!}(\vt z){\vt d}$ in another space-fixed frame $\mathcal S_z$. It follows that $\vt \xi = \mt A(\vt g)\vt \delta$, where the composite rotation matrix $\mt A(\vt g) = \tr{\mt A\!}(\vt z){\mt A}(\vt q)$ denotes the relative orientation between $\mathcal S_q$ and $\mathcal S_z$. One key result in quaternion algebra\cite{Dichmann1999} is a simple relation that exists between $\vt q$, $\vt z$, and $\vt g$, which can be adapted to our notation as $\vt g = \tr{[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}]} \vt q$. Note that both $\vt q$ and $\vt z$ vary while the body rotates, but the quaternion $\vt g$ remains constant because $\mathcal S_q$ and $\mathcal S_z$ are fixed in space. Moreover, as Eq.~\ref{eq:relation_qdot_omega} was derived for an arbitrary space-fixed frame, it is also true that $\dot{\vt z} = \tfrac{1}{2}{\vt B}(\vt z)\vt \omega$. In conclusion, as an alternative to tracking the evolution of $\vt q(t)$ after an instant $t_0$ in which $\vt q = \vt q_0$, one can select some other convenient unit quaternion $\vt z_0$ and then track the evolution of $\vt z(t)$, starting from $\vt z(t_0) = \vt z_0$. Then, whenever $\vt q$ is needed, one can simply make 
\begin{equation}
\label{eq:relation_q_z}
\vt q = [\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}] \tr{[\begin{array}{cc}\vt z_0 & \mt C(\vt z_0)\end{array}]} \vt q_0,
\end{equation}
which is correct because these block matrices represent orthogonal transformations in $\mathbb{R}^4$.

@@

The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) is used here to develop a Hamiltonian formalism for the motion of a rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations. We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively. Since $K = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + \frac{1}{2} \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
\begin{equation*}
\mathcal{L} = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + 2 \tr{\dot{\vt q}} \mt B \mt I \tr{\mt B} \dot{\vt q} - U(\vt r, \vt q).
\end{equation*}

To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\cite{Goldstein2002} By carrying out the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
\begin{equation}
\label{eq:conj_momentum}
\vt \pi = 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega.
\end{equation}

Hence, $\vt \pi$ is twice the quaternion-fixed version of the angular momentum $\vt L = \mt I \vt \omega$ and, as such, it is also orthogonal to $\vt q$. Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
\begin{equation}
\label{eq:H_with_B}
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + \frac{\tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi}{8} + U(\vt r, \vt q).
\end{equation}

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} Eq.~\ref{eq:H_with_B} is convenient for differentiation with respect to $\vt \pi$, since $\mt B(\vt q)$ does not depend on it. However, if we define $\mt \Omega = \mt B(\vt \pi)$, it follows that $\tr{\mt B}{\vt \pi} = - \tr{\mt \Omega}{\vt q}$ and, therefore, a convenient form of $\mathcal{H}$ for differentiation with respect to $\vt q$ is
\begin{equation}
\label{eq:H_with_Omega}
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + \frac{\tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q}{8} + U(\vt r, \vt q).
\end{equation}

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame. Therefore, $-\partial U/\partial \vt q$ is twice the torque represented in quaternion-fixed frame of reference. Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = - \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \label{eq:EDO_q} \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:EDO_pi}
\end{align}
\end{subequations}

@@

\subsection{Exact Solution for Free Rotation}
\label{sec:free_rotation}

Eq.~\ref{eq:EDO_system} cannot be integrated analytically for a general potential $U(\vt r,\vt q)$, but closed-form solutions for Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi} are known in the case when $\vt \tau = \vt 0$. Torque-free rotations are useful in Molecular Dynamics as part of numerical integration methods based on splitting strategies. This is the topic addressed in Sec.~\ref{sec:symplectic}. As shown in Appendix~\ref{sec:derivation_euler}, Eq.~\ref{eq:EDO_pi} results in the well-known Euler equation $\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau$, which is totally decoupled from Eq.~\ref{eq:EDO_q} when $\vt \tau = \vt 0$. It is a known fact that both the rotational energy $K_r = \frac{1}{2}\tr{\vt \omega}{\mt I}{\vt \omega}$ and the space-fixed angular momentum vector $\vt L$ (whose squared norm is $L^2 = \tr{\vt \omega}{\mt I}^2{\vt \omega}$) are integrals of motion in a free rotation. This allows us to solve the Euler equation analytically for $\vt \omega$. In the case of a spherical top ($I_1 = I_2 = I_3$), the solution is simply $\vt \omega(t) = \vt \omega(0)$. Inspired by the derivations in Refs.~\onlinecite{Landau1976, Thorn2012, vanZon2007}, we present the general solution in a form that facilitates computer implementation and allows a unified treatment for both asymmetric ($I_1 > I_2 > I_3$) and symmetric tops (either $I_1 = I_2$ or $I_2 = I_3$) if infinities are properly dealt with (IEEE 754 standard). From an initial condition $\vt \omega(0) = \vt \omega_0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, the solution starts by computing $K_r = \frac{1}{2}\tr{\vt \omega_0}{\mt I}{\vt \omega}_0$ and $L = (\tr{\vt \omega_0}{\mt I}^2{\vt \omega_0})^{1/2}$, followed by evaluating a set of prefactors defined as
\begin{align*}
a_1 &= \text{sgn}(\omega_1^0)\sqrt{\frac{L^2 - 2 I_3 K_r}{I_1(I_1 - I_3)}}, \\
a_2 &= \sqrt{\min(\lambda_1,\lambda_3)}, \quad \text{and} \\
a_3 &= \text{sgn}(\omega_3^0)\sqrt{\frac{2 I_1 K_r - L^2}{I_1(I_1 - I_3)}}, \\
\end{align*}
where $\lambda_1 = \frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}$ and $\lambda_3 = \frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}$. Note that the value of $a_2$ depends on a comparison between $\lambda_1$ and $\lambda_3$. This also occurs for $\vt \omega$, whose functional form is
\begin{subequations}
\begin{align}
\vt \omega(t) &= \begin{smallarray}{c}
a_1\text{cn}(\omega_p t + \epsilon) \\
a_2\text{sn}(\omega_p t + \epsilon) \\
a_3\text{dn}(\omega_p t + \epsilon)
\end{smallarray} \quad \text{if} \quad \lambda_1 < \lambda_3 \quad \text{or} \\
\vt \omega(t) &= \begin{smallarray}{c}
a_1\text{dn}(\omega_p t + \epsilon) \\
a_2\text{sn}(\omega_p t + \epsilon) \\
a_3\text{cn}(\omega_p t + \epsilon)
\end{smallarray} \quad \text{if} \quad \lambda_1 > \lambda_3,
\end{align}
\end{subequations}
where $\text{sn}$, $\text{cn}$, and $\text{dn}$ are the Jacobi elliptic functions\cite{Byrd_1971} with parameter $m = k^2 = {a_2}/{\text{max}(\lambda_1,\lambda_3)}$ and the constants $\omega_p$ and $\epsilon$ are given by
\begin{equation*}
\omega_p = -\frac{(I_1 - I_3)a_1 a_3}{I_2 a_2} \quad \text{and} \quad
\epsilon = \text{sn}^{-1}\left(\frac{\omega_2^0}{a_2}\right).
\end{equation*}

In order to compute the inverse function $\text{sn}^{-1}$, one can resort to its relation with an elliptic integral in Carlson form,\cite{Carlson_1977, Gradshteyn2014} $\text{sn}^{-1}(x) = x R_F(1-x^2,1-mx^2,1)$. One can numerically evaluate Jacobi functions and Carlson integrals by using, for instance, the routines available in the GNU Scientific Library.\cite{Galassi_2009}

Once $\vt \omega(t)$ is known, we can solve Eq.~\ref{eq:EDO_q} for the unit quaternion $\vt q$ with initial condition $\vt q(0) = \vt q_0$. Here we propose a new procedure that is closely linked to the methods by Kosenko\cite{Kosenko_1998} and by Celledoni \textit{et al}.,\cite{Celledoni_2008} but that relies on the rotation matrix factorization (Eq.~\ref{eq:factorization_of_A}) to facilitate the task. These methods are also related to the one by van Zon and Schofield,\cite{vanZon2007} as shown in Ref.~\onlinecite{Celledoni_2008}. During a free rotation, any space-fixed representation of the angular momentum vector will remain constant. Thus, instead of solving Eq.~\ref{eq:EDO_q} directly for obtaining $\vt q(t)$, we solve an analogous equation $\dot{\vt z} = \frac{1}{2}{\mt B(\vt z)}{\vt \omega}$ for determining a unit quaternion $\vt z(t)$ specially chosen so that $\tr{\mt A\!}(\vt z)\mt I\vt \omega = \tr{[\begin{array}{ccc} L & 0 & 0 \end{array}]}$. In other words, we look for an alternative space-fixed frame whose first axis is aligned with the angular momentum vector. After replacing $\tr{\mt A\!}(\vt z) = \tr{\mt C}(\vt z){\mt B}(\vt z)$, we premultiply both sides by $\mt C(\vt z)$ and then apply Eq.~\ref{eq:product_B_vector} to obtain the homogeneous linear system $\mt G \vt z = \vt 0$, where $\mt G = L \hat{\vt C}_1 - \sum_{j=1}^3 I_j \omega_j \hat{\vt B}_j$. This means that $\vt z$ will continuously lie in the kernel of matrix $\mt G$, which can be determined by a standard elimination method. Such kernel turns out to be a (moving) linear space defined by the orthogonal basis vectors
\begin{equation*}
\vt u(t) = \begin{smallarray}{c} I_3\omega_3(t) \\ I_2\omega_2(t) \\ L-I_1\omega_1(t) \\ 0 \end{smallarray}
\quad \text{and} \quad
\vt v(t) = \begin{smallarray}{c} -I_2\omega_2(t) \\ I_3\omega_3(t) \\ 0 \\ L-I_1\omega_1(t) \end{smallarray}.
\end{equation*}

At all times, therefore, $\vt z$ will be a linear combination of $\vt u$ and $\vt v$. We can then write the expression
\begin{equation*}
\vt z(t) = \frac{b(t) \vt u(t) + c(t) \vt v(t)}{\sqrt{2L[L-I_1\omega_1(t)]}},
\end{equation*}
whose denominator is the norm of both $\vt u$ and $\vt v$. The problem has now been reduced to determining how the coefficients $b$ and $c$ vary with time. Further reduction is achieved by realizing that both the unit-norm constraint ($\tr{\vt z}{\vt z} = 1$) and the orthogonality condition ($\tr{\vt u}{\vt v} = 0$) hold at all times, which makes
\begin{equation*}
\tr{\vt z}{\vt z} = \frac{b^2 \tr{\vt u}{\vt u} + 2bc\tr{\vt u}{\vt v} + c^2 \tr{\vt v}{\vt v}}{2L[L-I_1\omega_1(t)]} = b^2 + c^2 = 1.
\end{equation*}

Hence, the values of $b$ and $c$ are restricted to a unit circle, so that we can finally rewrite $\vt z(t)$ as
\begin{equation*}
\vt z(t) = \frac{\vt u(t) \cos \theta(t) + \vt v(t) \sin \theta(t)}{\sqrt{2L[L-I_1\omega_1(t)]}}.
\end{equation*}

The next step consists in replacing this result into the equation $\dot{\vt z} = \frac{1}{2}{\mt B(\vt z)}{\vt \omega}$. This yields, after some algebra, a single differential equation whose solution can be expressed as
\begin{equation*}
\theta(t) = \theta_0 + \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2L I_1}\int_0^t \frac{1}{1-(I_1/L)\omega_1(s)}ds.
\end{equation*}

In order to solve this equation, we introduce the following functions derived from definite integrals: $C(x) = \int_0^x \frac{1}{1 - \alpha \cn y}dy$ and $D(x) = \int_0^x \frac{1}{1 - \alpha \dn y}dy$, where $\alpha = I_1a_1/L$.

\begin{widetext}
	\begin{subequations}
\begin{gather}
\label{eq:def_integral_C} \frac{C(x)}{\eta + 1} = x + \text{sgn}(\cn x) \Theta(\sn x,\eta) + 2\Theta(1,\eta) \left\lfloor \frac{x}{2 K} \right\rceil +
\frac{\alpha}{\sqrt{m+\eta}} \arctan\left(\frac{\sqrt{m+\eta} \; \sn x}{\dn x}\right) \\
\label{eq:def_integral_D} \frac{D(x)}{\eta + 1} = x + \text{sgn}(\cn x) \Theta(\sn x,m\eta) + 2\Theta(1,m\eta) \left\lfloor\frac{x}{2 K}\right\rceil +
\frac{\alpha}{\sqrt{m\eta + 1}} \left\{\arctan\left(\frac{\sqrt{m\eta + 1} \; \sn x}{\cn x}\right) + \pi \left\lfloor \frac{x}{2 K} \right\rceil\right\}
\end{gather}
	\end{subequations}
\end{widetext}

\begin{figure}
	\includegraphics{definite_integrals.eps}
	\caption{Illustrative plot of functions $c(x)=1/(1-\alpha\cn x)$ (dark cyan line) and $d(x)=1/(1-\alpha \dn x)$ (magenta line) with $\alpha = 9/10$ and elliptic parameter $m = 1/2$, together with their definite integrals $C(x)$ (blue line) and $D(x)$ (red line), defined in Eqs.~\ref{eq:def_integral_C} and \ref{eq:def_integral_D}, respectively.}
	\label{fig:definite_integrals}
\end{figure}

@@

An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element. Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
\begin{equation*}
\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
\end{equation*}
where $i\!L$ is the Liouville operator. This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$. As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$. Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.

Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}. Let us analyze three of them, which will be useful later on to derive a complete approximate solution. In what follows, an index $0$ stands for ``initial value''. We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$. The exact solution for this case is
\[
{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0,
\]
with ${\vt p}(t) = {\vt p}_0$, ${\vt q}(t) = {\vt q}_0$, and ${\vt \pi}(t) = {\vt \pi}_0$. This corresponds to a free translation (hence the subscript $t$). As the momentum $\vt p$ remains constant, the Hamiltonian $\mathcal{H}_t$ is conserved, as expected. Of course, the solution above is the exact effect of the propagator $e^{t i\!L_t}$, where the Liouville operator $i\!L_t$ is, simply,
\[
i L_t = \frac{\tr{\vt p}}{m}\diff{}{\vt r}.
\]

In the second hypothetical case, referred to as the boost case, we consider a Hamiltonian $\mathcal{H}_b = U(\vt r, \vt q)$. The resulting equations of motion are $\dot{\vt r} = 0$, $\dot{\vt p} = \vt F$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 2 \mt C \vt \tau$, whose exact solution is
\begin{align*}
&{\vt p}(t) = {\vt p}_0 + t \vt F \\
&{\vt \pi}(t) = {\vt \pi}_0 +  2 t \mt C \vt \tau
\end{align*}
with ${\vt r}(t) = {\vt r}_0$ and ${\vt q}(t) = {\vt q}_0$. As $\vt r$ and $\vt q$ remain unaltered, $\vt F$, $\mt C$, and $\vt \tau$ are constant as well. This solution clearly conserves $\mathcal{H}_b$ and marks the effect of a propagator $e^{t i\!L_b}$, obtained from the operator
\[
i\!L_b = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C} \diff{}{\vt \pi}.
\]

Finally, the third hypothetical case consists of a Hamiltonian $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Note that a single entry of $\vt \omega$ is accounted for in this case, so that it corresponds to a free uniaxial rotation. The equations of motion resulting from $\mathcal{H}_k$ are $\dot{\vt r} = 0$, $\dot{\vt p} = 0$, $\dot{\vt q} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt q$, and $\dot{\vt \pi} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt \pi$. A special feature of this system facilitates its solution. It is known beforehand that the solution will keep $\mathcal{H}_k$ unaltered and, as a consequence, $\omega_k$ will be a constant as well. As noticed by Miller~\textit{el al.}~\cite{Miller2002}, this decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form. The analytical solution that takes place is\cite{Miller2002}
\begin{align*}
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
\end{align*}
with ${\vt r}(t) = {\vt r}_0$ and ${\vt p}(t) = {\vt p}_0$. These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{t i\!L_k}$ derived from the operator
\[
i\!L_k = \frac{1}{2} \omega_k \left( \tr{\vt q}\tr{\hat{\mt B}_k} \diff{}{\vt q} + \tr{\vt \pi}\tr{\hat{\mt B}_k} \diff{}{\vt \pi} \right).
\]

>>>>

\section{New Theory}
Exact solution for torque-free rotation
In a previous paper\cite{Abreu_2017} we presented a formulation for the motion of rigid bodies which was simplified by the introduction of a particular factorization of the rotational matrix. 
quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ and satisfy the relation\cite{Goldstein2002}
\begin{equation}
\label{eq:norm_eq_1}
\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1.
\end{equation}

Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray}.
\end{equation*}

The most remarkable feature is that
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation}

Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$.

\section{Introduction}
\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}
\cite{Omelyan_2011} multiple time scale md for fluids with orientational dof. canonical and isokinetic ensembles.
Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations. This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.

@@

\subsection{Kinematics}

@@

The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) is used here to develop a Hamiltonian formalism for the motion of a rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \label{eq:EDO_q} \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:EDO_pi}
\end{align}
\end{subequations}

@@

@@

An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element. 
\subsection{Exact Solution for Free Rotation}
\label{sec:free_rotation}
